Understanding the Lomb-Scargle Periodogram

### 2017-02-06 PracticalLombScargle.tex
add to post-script discussion

\subsubsection{False Alarm Probability}
A much more relevant quantity for expressing uncertainty of periodogram results
is the height of the peak, and in particular the height compared to the
background ``noise'' displayed in the periodogram.
\fig{peak-width-height} indicates that this property {\it does} depend on
both the number of observations and their signal-to-noise ratio: for the
small-$N$/low signal-to-noise cases, the spurious peaks in the background
become comparable to the size of the true peak.
In fact, as we saw in \sect{schuster-to-lomb-scargle}, the ability to
analytically define and quantify the relationship between peak height
and significance is one of the primary considerations that led to the
standard form of the Lomb-Scargle periodogram.

@@

\todo{Summarize key recommendations}

\begin{itemize}
\item always use the floating-mean model---it has many benefits with virtually zero cost
\item always explore the window power first, so that you know what alias structures to expect
\item the peak of the periodogram rarely tells the entire story. Always check
  for aliases that are expected from window structure ($f = f_0 \pm n\delta f$)
  and from non-sinusoidal signals ($f = f_0 / m$).
\item avoid Bayesian methods unless you are very comfortable with the
  assumptions they make about your signal
\item avoid peak widths as a proxy for uncertainty unless you are quite
  certain that your analysis has found the correct peak.
\item avoid false alarm probabilities as a proxy for uncertainty
  unless you make very clear {\it exactly} what kind of error they measure.
  Estimate the FAP via bootstrap if computationally feasible, and otherwise
  keep in mind the failures of available estimates.
\end{itemize}

\subsection{Postscript: Why Lomb-Scargle?}
After considering all of these practical aspects, I think it is worth
revisiting the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.
In particular, after three decades of use, there seems to be an unrealistic
mythology about the efficacy of the Lomb-Scargle approach among astronomers.
It's common to see statements or implications like the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations})
  \item {\it Myth: The Lomb-Scargle periodogram is faster than the least squares
    periodogram because it avoids explicitly solving for model coefficients.}
    Reality: model coefficients can be determined with not much extra
    computation \citep[see discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    peak-value statistics, which must be either approximated or computed by
    bootstrap resampling.
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same partial aliasing issues as the Schuster
    periodogram (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: it rather
    shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding a generic, non-sinusoidal signal.
\end{itemize}
With these myths debunked, what advantage remains for Lomb-Scargle?
What advantages would we lose by instead using a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?
I think we only lose the advantage of being able to generalize to more
complicated models in the Bayesian sense, in which case you're not really
using Lomb-Scargle at all, but rather a generative Bayesian model for your
data.
The point of correspondance for a sinusoidal model is perhaps useful trivia,
but not a reason to use that model if your data is not known to be sinusoidal.

Conversely, if you use a Schuster approach rather than a Lomb-Scargle approach,
what you gain is the ability to reason quantitatively about the effects of the
survey window function on the resulting periodogram (see \sect{CLEAN}).
While the deconvolution problem is well-known to be ill-posed, there may be a
way forward for cleaned periodograms based on regularized deconvolutions.

>>>>

\subsubsection{False Alarm Probability}
\sectlabel{false-alarm-probability}

A much more relevant quantity for expressing uncertainty of periodogram results
is the height of the peak, and in particular the height compared to the
background ``noise'' displayed in the periodogram.
\fig{peak-width-height} indicates that this property {\it does} depend on
both the number of observations and their signal-to-noise ratio: for the
small-$N$/low signal-to-noise cases, the spurious peaks in the background
become comparable to the size of the true peak.
In fact, as we saw in \sect{schuster-to-lomb-scargle}, the ability to
analytically define and quantify the relationship between peak height
and significance is one of the primary considerations that led to the
standard form of the Lomb-Scargle periodogram.

@@

\todo{Flesh these out}

\begin{itemize}
\item choose your frequency range \& grid spacing carefully, so as to not
  miss relevant peaks.
\item always compute the window periodogram first, so that you know what alias
  structures to expect in the periodogram for your data.
\item always use the floating-mean model---it has many benefits with
  virtually zero down-sides.
\item the highest peak of the periodogram rarely tells the entire story.
  Always check for aliases that are expected from window structure
  ($f = f_{max} \pm n\delta f$) and from non-sinusoidal signals
  ($f = f_{max} / m$).
\item avoid Bayesian methods unless you are very comfortable with the
  assumptions they make about the form of your signal.
\item avoid peak widths as a proxy for uncertainty unless you are quite
  certain that your analysis has found the correct peak.
\item avoid false alarm probabilities as a proxy for uncertainty
  unless you make very clear {\it exactly} what kind of error they measure.
  Estimate the FAP via bootstrap if computationally feasible, and otherwise
  keep in mind the failures of available estimates.
\item in the context of a time-domain survey, always use simulated observations
  to evaluate your rates of false positives, false negatives, and failures to
  identify the correct frequency.
\item in cases of strong aliasing, consider using Lomb-Scargle as a first pass,
  and then select among candidate frequencies with a more detailed model of
  the signals you expect.
\end{itemize}

\subsection{Postscript: Why Lomb-Scargle?}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.
In particular, three decades after the entry of ``Lomb-Scargle'' into the
astronomy community's vernacular, there seems to be a well-developed
mythology surrounding the efficacy and efficiency of the Lomb-Scargle approach.
It's common to see statements or implications like the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with not much extra
    computation \citep[see discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    peak-value statistics, which must be either approximated or computed by
    bootstrap resampling (see \sect{false-alarm-probability})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same partial aliasing issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: it
    shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding a generic, non-sinusoidal signal
    (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths debunked, what advantage remains for Lomb-Scargle?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

I think we only lose the advantage of being able to generalize to more
complicated models in the Bayesian sense, in which case you're not really
using Lomb-Scargle at all, but rather a generative Bayesian model for your
data based on some strong prior information about the form of your signal.
The equivalence of Lomb-Scargle to a Bayesian sinusoidal model is perhaps
interesting trivia, but not itself a reason to use that model if your data
is not known {\it a priori} to be sinusoidal.

Conversely, if you replace your Lomb-Scargle approach with a Schuster approach,
what you gain is the ability to reason quantitatively about the effects of the
survey window function on the resulting periodogram (see \sect{CLEAN}).
While the deconvolution problem is ill-posed, there is no reason to assume
this is a fatal defect: ill-posed linear models are solved routinely in
other areas of computational research, particularly by using priors or
various forms of regularization.

One clear advantage of both Lomb-Scargle and Schuster periodograms is that
they are quite fast to compute compared to more sophisticated approaches to
periodic analysis based on more constrained generative models or Bayesian
non-parametric models, and in the context of large time-domain surveys can
provide rough periodicity detections and catalog parameters in near real-time.
